---
title: "Genomic Prediction Analysis - Stage II of II: Cross-validation Round 2"
author: "wolfemd"
date: "2019-7-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

# Objective

This time with the outliers-removed BLUPs. Based on results in round 1, did not continue with some of the traits.

# Set-up training data
```{r, eval=F}
rm(list=ls()); gc()
library(tidyverse); library(magrittr); 
blups<-readRDS(file="data/iita_blupsForCrossVal_outliersRemoved_73019.rds")
K<-readRDS(file=paste0("/workdir/IITA_2019GS/Kinship_IITA_TrainingPop_72619.rds"))
blups %<>%
      rename(trainingData=blups) %>% 
      mutate(trainingData=map(trainingData,~filter(.,GID %in% rownames(K))),)
tms13f<-rownames(K) %>% grep("TMS13F|2013_",.,value = T); length(tms13f) # 2395
tms14f<-rownames(K) %>% grep("TMS14F",.,value = T); length(tms14f) # 2171
tms15f<-rownames(K) %>% grep("TMS15F",.,value = T); length(tms15f) # 835
gg<-setdiff(rownames(K),c(tms13f,tms14f,tms15f)); length(gg) # 1228 (not strictly gg)

blups %<>%
    mutate(seed_of_seeds=1:n(),
           seeds=map(seed_of_seeds,function(seed_of_seeds,reps=5){ 
               set.seed(seed_of_seeds); 
               outSeeds<-sample(1:1000,size = reps,replace = F); 
               return(outSeeds) }))
blups %<>%
      select(-varcomp); gc()
```

# Cross-validation function 
```{r, eval=F}
# trainingData<-blups$trainingData[[1]]; seeds<-blups$seeds[[1]]; nfolds<-5; reps<-5;
crossValidateFunc<-function(Trait,trainingData,seeds,nfolds=5,reps=5,ncores=50,...){
      trntstdata<-trainingData %>% 
            filter(GID %in% rownames(K))
      K1<-K[rownames(K) %in% trntstdata$GID,
            rownames(K) %in% trntstdata$GID]
      # rm(K,trainingData); gc()
      # seed<-seeds[[1]]
      # Nfolds=nfolds
      makeFolds<-function(Nfolds=nfolds,seed){
            genotypes<-rownames(K1)
            
            set.seed(seed)
            seed_per_group<-sample(1:10000,size = 4,replace = FALSE)
      
            set.seed(seed_per_group[1])
            FoldsThisRep_tms15<-tibble(CLONE=genotypes[genotypes %in% tms15f],
                                       Group="TMS15F") %>% 
                  mutate(Folds=sample(1:Nfolds,nrow(.),replace=T)) %>% 
                  arrange(Folds) %>% 
                  group_by(Group,Folds) %>% 
                  nest(.key = Test)
            set.seed(seed_per_group[2])
            FoldsThisRep_tms14<-tibble(CLONE=genotypes[genotypes %in% tms14f],
                                       Group="TMS14F") %>% 
                  mutate(Folds=sample(1:Nfolds,nrow(.),replace=T)) %>% 
                  arrange(Folds) %>% 
                  group_by(Group,Folds) %>% 
                  nest(.key = Test)
            set.seed(seed_per_group[3])
            FoldsThisRep_tms13<-tibble(CLONE=genotypes[genotypes %in% tms13f],
                                       Group="TMS13F") %>% 
                  mutate(Folds=sample(1:Nfolds,nrow(.),replace=T)) %>% 
                  arrange(Folds) %>% 
                  group_by(Group,Folds) %>% 
                  nest(.key = Test)
            set.seed(seed_per_group[4])
            FoldsThisRep_gg<-tibble(CLONE=genotypes[genotypes %in% gg],
                                       Group="GGetc") %>% 
                  mutate(Folds=sample(1:Nfolds,nrow(.),replace=T)) %>% 
                  arrange(Folds) %>% 
                  group_by(Group,Folds) %>% 
                  nest(.key = Test)
           FoldsThisRep<-bind_rows(FoldsThisRep_tms15,FoldsThisRep_tms14) %>% 
                  bind_rows(FoldsThisRep_tms13) %>% 
                  bind_rows(FoldsThisRep_gg) %>% 
                  mutate(Test=map(Test,~.$CLONE),
                         Train=map(Test,~genotypes[!genotypes %in% .]))
            return(FoldsThisRep) }
      
    crossval<-tibble(Rep=1:reps,seed=unlist(seeds)[1:reps]) %>% 
        mutate(Folds=map2(Rep,seed,~makeFolds(Nfolds=nfolds,seed=.y))) %>% 
        unnest()

    #Test<-crossval$Test[[1]]; Train<-crossval$Train[[1]]
    crossValidate<-function(Test,Train){ 
      train<-Train
      test<-Test
      trainingdata<-trntstdata %>% 
        filter(GID %in% train) %>% 
        mutate(GID=factor(GID,levels=rownames(K1)))
       
      require(sommer)
      proctime<-proc.time()
      fit <- mmer(fixed = drgBLUP ~1,
                  random = ~vs(GID,Gu=K1),
                  weights = WT,
                  data=trainingdata) 
      proc.time()-proctime
      
      x<-fit$U$`u:GID`$drgBLUP 
      gebvs<-tibble(GID=names(x),
                    GEBV=as.numeric(x))
      
      accuracy<-gebvs %>% 
            filter(GID %in% test) %>% 
            left_join(
                  trntstdata %>% 
                        dplyr::select(GID,BLUP) %>% 
                        filter(GID %in% test)) %$% 
            cor(GEBV,BLUP, use='complete.obs') 
    return(accuracy)
    }
    
    require(furrr)
    options(mc.cores=ncores)
    plan(multiprocess)
    crossval<-crossval %>% 
          mutate(accuracy=future_map2(Test,Train,~crossValidate(Test=.x,Train=.y)))
    saveRDS(crossval,file=paste0("/workdir/IITA_2019GS/CrossVal_73019/",
                                 "CrossVal_",Trait,"_OutliersRemoved_73019.rds"))
    rm(list=ls()); gc() 
    }
```
# Run CV on two servers

# cbsulm14 (112)
```{r, eval=F}
blups %>% 
      mutate(CVaccuracy=pmap(.,crossValidateFunc))
#saveRDS(cvresults_1,file="/workdir/IITA_2019GS/CrossValResults_IITA_TrainingPop_1_72719.rds")
```

# Results
```{r}
rm(list=ls());gc()
library(tidyverse); library(magrittr); library(cowplot); 
cvNoOutliers<-tibble(Files=list.files("output/CrossVal_73019/")) %>% 
                  mutate(Trait=gsub("CrossVal_","",Files),
                         Trait=gsub("_OutliersRemoved_73019.rds","",Trait),
                         Dataset="OutliersRemoved") %>% 
                  mutate(cvResults=map(Files,~readRDS(paste0("output/CrossVal_73019/",.)))) %>% 
      dplyr::select(-Files)
cvWithOutliers<-tibble(Files=list.files("output/CrossVal_72719/")) %>% 
      filter(grepl("HistoricalDataIncluded|BRNHT1|PLTHT",Files)) %>% 
      mutate(Trait=gsub("CrossVal_","",Files),
             Trait=gsub("_2013toPresent_72719.rds","",Trait),
             Trait=gsub("_HistoricalDataIncluded_72719.rds","",Trait),
             Dataset="NoOutlierRemoval") %>% 
      filter(Trait %in% cvNoOutliers$Trait) %>% 
      mutate(cvResults=map(Files,~readRDS(paste0("output/CrossVal_72719/",.)))) %>% 
      dplyr::select(-Files)
cv<-bind_rows(cvNoOutliers,
              cvWithOutliers)

cv %<>% 
  unnest(cols = cvResults) %>% 
  mutate(Ntrain=map_dbl(Train,length),
         Ntest=map_dbl(Test,length)) %>% 
  select(-Test,-Train) %>% 
  unnest(cols = accuracy)
```

## Figure 1

I did an additional cross-validation, using BLUPs produced after two rounds of model-fitting, followed-by outlier removal. I defined outliers as observations with abs(studentized residuals)>3.3.
Overall, the improvement is not consistent or large, but I’d probably trend towards using the data with outliers removed.

By genetic group
```{r, fig.width=9, fig.height=6}
library(viridis)
cv %>% 
      ggplot(.,aes(x=Trait,y=accuracy,fill=Dataset)) + 
      geom_boxplot() + 
      facet_grid(.~Group,space='free_x',scale='free_x') +
      geom_hline(yintercept = 0,color='darkred',size=1.25) + 
      theme_bw() + 
      theme(axis.text.x = element_text(angle=90,face='bold',size=14)) +
      scale_fill_viridis_d()
```

## Figure 2

overall
```{r, fig.width=9, fig.height=6}
library(viridis)
cv %>% 
      ggplot(.,aes(x=Trait,y=accuracy,fill=Dataset)) + 
      geom_boxplot() + 
      geom_hline(yintercept = 0,color='darkred',size=1.25) + 
      theme_bw() + 
      theme(axis.text.x = element_text(angle=90,face='bold',size=14)) +
      scale_fill_viridis_d()
```

# Next step

[Stage II: Cross-validation Run 2](IITA_StageII_Predict_C4.html)
