---
title: "Predict GEBV for IITA GS C4"
author: "wolfemd"
date: "2019-7-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

# Prep. data for genomic prediction

## Load snps
```{r, eval=F}
rm(list=ls()); gc()
library(tidyverse); library(magrittr); 
snps<-readRDS(file=paste0("/workdir/ImputationStageIII_72619/",
                          "DosageMatrix_RefPanelAndGSprogeny_ReadyForGP_73019.rds"))
```
## Load blups
```{r, eval=F}
blupsNoOut<-readRDS(file="data/iita_blupsForCrossVal_outliersRemoved_73019.rds")
blups<-readRDS(file="data/iita_blupsForCrossVal_72619.rds")

blups<-blupsNoOut %>% 
  dplyr::select(Trait,blups) %>% 
  mutate(OutliersRemoved=TRUE) %>% 
  bind_rows(
    blups %>% 
      filter(Dataset=="HistoricalDataIncluded" | Trait %in% c("BRNHT1","PLTHT")) %>% 
      dplyr::select(Trait,blups) %>% 
      mutate(OutliersRemoved=FALSE))
```
# Match TMS18F germNames to DArT genos
For the TMS18F, there are some phenotypes available now. 
```{r, eval=F}
gidsInPhenos<-tibble(GID=blups %>% unnest() %$% GID %>% unique)
table(gidsInPhenos$GID %in% rownames(snps)) # 6629 (that only includes GG through C3)

# gidsInPhenos$GID %>% grep("TMS18",.,value=T) %>% length # 1452
# rownames(snps) %>% grep("TMS18",.,value=T) %>% length # 2420
gidsInPhenos %<>% 
  left_join(
    tibble(dartName=rownames(snps) %>% grep("TMS18",.,value=T)) %>% 
      separate(dartName,c("GID","dartID"),"_",extra = 'merge',remove = F) %>% 
      semi_join(gidsInPhenos) %>% 
      group_by(GID) %>% 
      slice(1) %>% 
      ungroup()) %>% 
  mutate(NewName=ifelse(!is.na(dartName),dartName,GID))

blups %<>% 
  unnest(blups) %>% 
  left_join(gidsInPhenos) %>% 
  mutate(GID=NewName) %>% 
  dplyr::select(-dartName,-dartID,-NewName) %>% 
  group_by(Trait,OutliersRemoved) %>% 
  nest(.key = "trainingData")
```

# Make a kinship matrix including C4

## Subset snps
```{r, eval=F}
table(gidsInPhenos$NewName %in% rownames(snps))
iitaSamples2keep<-union(rownames(snps) %>% .[. %in% gidsInPhenos$NewName],
                        rownames(snps) %>% grep("TMS18",.,value = T)) #%>% length # 9049
snps<-snps[iitaSamples2keep,]
```

## MAF>1% filter
```{r, eval=F}
maf_filter<-function(snps,thresh){
  freq<-colMeans(snps, na.rm=T)/2; maf<-freq;
  maf[which(maf > 0.5)]<-1-maf[which(maf > 0.5)]
  snps1<-snps[,which(maf>thresh)];
  return(snps1) }
snps %<>% maf_filter(.,0.01)
dim(snps) # [1]  9049 68029
```

## Make kinships
```{r, eval=F}
library(tidyverse); library(magrittr); library(rrBLUP)
K<-rrBLUP::A.mat(snps-1)

saveRDS(K,file=paste0("/workdir/marnin/IITA_2019GS/",
                      "Kinship_IITA_SamplesForGP_73119.rds"))
saveRDS(snps,file=paste0("/workdir/marnin/IITA_2019GS/",
                         "DosageMatrix_IITA_SamplesForGP_73119.rds"))
rm(snps); gc()
```


# PCA 

On cbsurobbins
```{r, eval=F}
pca<-prcomp(snps, scale=T, center=T) 
pc_scores<-pca$x %>% 
  .[,1:15] %>% 
  as.data.frame %>% 
  rownames_to_column(var="FullSampleName") %>% 
  mutate(GeneticGroup=NA,
         GeneticGroup=ifelse(grepl("TMS13F|2013_",FullSampleName),"TMS13F",
                             ifelse(grepl("TMS14F",FullSampleName),"TMS14F",
                                    ifelse(grepl("TMS15F",FullSampleName),"TMS15F",
                                           ifelse(grepl("TMS18F",FullSampleName),"TMS18F","GGetc")))))
```
```{r, eval=F}
summary(pca)$importance[,1:10] 
```
First two PCs _only_ explain about 9% of the genetic variance. 

Save PCA results
```{r, eval=F}
saveRDS(pc_scores,file="output/PCscores_IITA_SamplesForGP_73119.rds")
saveRDS(pca,file="/workdir/marnin/IITA_2019GS/PCA_IITA_SamplesForGP_73119.rds")
rm(pca); gc()
```


```{r}
rm(list=ls());gc()
library(tidyverse); library(magrittr); library(cowplot); 
pc_scores<-readRDS("output/PCscores_IITA_SamplesForGP_73119.rds")
library(viridis)
pc1v2<-pc_scores %>% 
  ggplot(.,aes(x=PC1,y=PC2,color=GeneticGroup)) + 
  geom_point() + 
  theme_bw() + scale_color_viridis_d() + 
  labs(x="PC1 (6%)",y="PC2 (3%)") + theme(legend.position = 'none')
pc3v4<-pc_scores %>% 
  ggplot(.,aes(x=PC3,y=PC4,color=GeneticGroup)) + 
  geom_point() + 
  theme_bw() + scale_color_viridis_d() + 
  labs(x="PC3 (3%)",y="PC4 (3%)")
plot_grid(pc1v2,pc3v4)
```

# Prediction 
On cbsulm16

## Load data for pred
```{r, eval=F}
rm(list=ls()); gc()
library(tidyverse); library(magrittr); 
K<-readRDS(file=paste0("/workdir/IITA_2019GS/Kinship_IITA_SamplesForGP_73119.rds"))
blupsNoOut<-readRDS(file="data/iita_blupsForCrossVal_outliersRemoved_73019.rds")
blups<-readRDS(file="data/iita_blupsForCrossVal_72619.rds")

blups<-blupsNoOut %>% 
  dplyr::select(Trait,blups) %>% 
  mutate(OutliersRemoved=TRUE) %>% 
  bind_rows(
    blups %>% 
      filter(Dataset=="HistoricalDataIncluded" | Trait %in% c("BRNHT1","PLTHT")) %>% 
      dplyr::select(Trait,blups) %>% 
      mutate(OutliersRemoved=FALSE))
gidsInPhenos<-tibble(GID=blups %>% unnest() %$% GID %>% unique)
gidsInPhenos %<>% 
  left_join(
    tibble(dartName=rownames(K) %>% grep("TMS18",.,value=T)) %>% 
      separate(dartName,c("GID","dartID"),"_",extra = 'merge',remove = F) %>% 
      semi_join(gidsInPhenos) %>% 
      group_by(GID) %>% 
      slice(1) %>% 
      ungroup()) %>% 
  mutate(NewName=ifelse(!is.na(dartName),dartName,GID))
blups %<>% 
  unnest(blups) %>% 
  left_join(gidsInPhenos) %>% 
  mutate(GID=NewName) %>% 
  dplyr::select(-dartName,-dartID,-NewName) %>% 
  filter(GID %in% rownames(K)) %>% 
  group_by(Trait,OutliersRemoved) %>% 
  nest(.key = "trainingData")
```

## Get GEBVs
```{r, eval=F}
require(sommer)
require(furrr)
options(future.globals.maxSize= 1500*1024^2)
options(mc.cores=18)
plan(multiprocess)
proctime<-proc.time()
blups %>% 
  mutate(GEBV=future_pmap(.,function(Trait,trainingData,OutliersRemoved,...){
    trndata<-trainingData %>% 
      mutate(GID=factor(GID,levels=rownames(K)))
    fit <- mmer(fixed = drgBLUP ~1,
                random = ~vs(GID,Gu=K),
                weights = WT,
                data=trndata)
    x<-fit$U$`u:GID`$drgBLUP 
    gebvs<-tibble(GID=names(x),
                  GEBV=as.numeric(x))
    saveRDS(gebvs,file=paste0("output/",
                              "GEBV_",Trait,"_OutliersEqual",OutliersRemoved,"_IITA_73119.rds"))
    rm(fit); gc()}))
proc.time() - proctime
```

## Format GEBV
```{r, eval=F}
gebvs<-list.files(path = "output/") %>% 
  grep("GEBV_",.,value = T) %>% 
  grep("_IITA_73119.rds",.,value = T) %>% 
  tibble(Files=paste0("output/",.)) %>% 
  mutate(Trait=gsub("output/GEBV_","",Files),
         Trait=gsub("_IITA_73119.rds","",Trait)) %>% 
  separate(Trait,c("Trait","OutliersRemoved"),"_OutliersEqual") %>% 
  mutate(Files=map(Files,~readRDS(.))) %>% 
  unnest(Files) %>% 
  mutate(GeneticGroup=NA,
         GeneticGroup=ifelse(grepl("TMS13F|2013_",GID),"TMS13F",
                             ifelse(grepl("TMS14F",GID),"TMS14F",
                                    ifelse(grepl("TMS15F",GID),"TMS15F",
                                           ifelse(grepl("TMS18F",GID),"TMS18F","GGetc")))))

```
## Write GEBVs
```{r, eval=F}
gebvs %>% 
  spread(Trait,GEBV) %>% 
  group_by(OutliersRemoved) %>% 
  nest() %>% 
  mutate(data=map2(data,OutliersRemoved,function(data,OutliersRemoved){
    write.csv(data,
              file=paste0("output/GEBV_IITA_OutliersRemoved",OutliersRemoved,"_73119.csv"),
              row.names = F) }))
```



